---
title: "Lab 1: Handling social network data in R"
#bibliography: references.bib
author: "Jola Gockel"
---

```{r, globalsettings, echo=FALSE, warning=FALSE, results='hide'}
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test2"))
options(width = 100)
rgl::setupKnitr()

colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

```{r setup, include=FALSE}
library(network)
library(igraph)
library(reshape2)
library(tidyverse)
```

Last compiled on `r format(Sys.time(), '%B, %Y')`

<br>

----
  
This is just an example page of your `r colorize("lab journal", "red")`.
  

----

# Getting network data into R: the igraph package

## From adjacency matrix to igraph object

```{r}
url1 <- "https://github.com/rensec/sasr08/raw/main/g_adj_matrix_simple.csv"
g_matrix <- read.csv(file = url1, header = FALSE)
g_matrix

class(g_matrix) #R automatically loads the adjacency matrix as a data frame

#turn data frame into matrix object
g_matrix <- as.matrix(g_matrix)
g_matrix

#adding row and column names for clarity
rownames(g_matrix) <- 1:nrow(g_matrix)
colnames(g_matrix) <- 1:ncol(g_matrix)
g_matrix

#importing file with attributes
g_nodes_age <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_nodes_age.csv")
g_nodes_age

#turn matrix into igraph network object
g <- graph_from_adjacency_matrix(g_matrix)
class(g)
g
plot(g)

#add node attributes to the network object
g <- set_vertex_attr(g, 
                     name = "age",
                     value = g_nodes_age$age)

```


## From edge list to igraph object

```{r}
g_elist <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_elist.csv")
g_elist

#creating igraph network object from edge list
g_from_elist <- graph_from_data_frame(g_elist) 
plot(g_from_elist) #shows: we lost the isolates

#to include isolates: also feed a list of all nodes
g_from_elist <- graph_from_data_frame(g_elist, vertices = g_nodes_age) 
plot(g_from_elist)
```

```{r}
#assignment: pipeline
g_elist %>%
  graph_from_data_frame(vertices = g_nodes_age) %>%
  plot()
```


## From adjacency list to igraph object

```{r}
g_adj_list <- read.csv(file = "https://github.com/rensec/sasr08/raw/main/g_adj_list.csv")
g_adj_list

#no way to immediately import adjacency lists to igraph network object
#instead: first transform adjacency list into edge list, then into igraph
#transform the data into long format for this
g_elist_from_alist <- g_adj_list
g_elist_from_alist <-  select(g_elist_from_alist, id, friend1:friend2) # keep only the network variables (and id)
g_elist_from_alist <-  melt(g_elist_from_alist, id.vars = "id") # from the reshape2 package
g_elist_from_alist

#same but as pipeline
g_elist_from_alist <- g_adj_list %>% 
  select(id, friend1:friend2) %>% #keep only the network variables (and id)
  melt( id.vars = "id") #from the reshape2 package

#same but using tidyverse
g_adj_list %>% # (we don't save it into an object as above this time, since we just want to show that the result is the same)
  select(id, friend1:friend2) %>% #keep only the network variables (and id)
  pivot_longer(c(friend1, friend2))

#to create edge list: drop all missing values on “value” and do some housekeeping
g_elist_from_alist <- g_elist_from_alist %>% 
  filter(!is.na(value)) %>% # drop the missings
  rename(from ="id", to = "value", sourcevar= "variable") %>% #just nice for interpretation
  relocate(to, .after=from) #move around the columns
g_elist_from_alist

#now import into igraph network object BUT we loose isolates
g_from_alist <- graph_from_data_frame(g_elist_from_alist) #this is a proper igraph graph object
plot(g_from_alist)

#instead: also feed the node list created from the adjacency list data frame
nodelist <- select(g_adj_list,id,age)
g_from_alist <- graph_from_data_frame(g_elist_from_alist, vertices = nodelist) #this is a proper igraph graph object
plot(g_from_alist)

```

```{r}
#assignment: pipeline
#g_adj_list TO BE ADDED
```


# The reverse direction: exporting from igraph objects

## From igraph to adjacency matrix

```{r}
g_adj_matrix <- as_adjacency_matrix(g, sparse = FALSE) #sparse=F to keep the zeros
g_adj_matrix

#check that this matrix matches the original file
all(g_adj_matrix == g_matrix)
```


## From igraph to edge list

```{r}
g_edgelist <- igraph::as_data_frame(g_from_elist, what = "edges") #:: to not use the dplyr/tidyverse function
g_edgelist 

#add nodes list to not loose isolates
igraph::as_data_frame(g_from_elist, what = "vertices") 
```


## From igraph to adjacency list

```{r}
edgelist_2 <- igraph::as_data_frame(g_from_alist, what = "edges")
nodelist_2 <- igraph::as_data_frame(g_from_alist, what = "vertices")

d <- edgelist_2 %>%
  rename(id = "from") %>% 
  pivot_wider( id_cols = id, names_from = sourcevar, values_from = to ) %>% 
  merge(nodelist_2, by.x = "id", by.y = "name", all.y = TRUE ) %>% 
  relocate(age, .after = id) %>% 
  type_convert() #as_data_frame returns characters; this transforms it back to numeric
d
```


# Getting network data into R: the network package

```{r}
#using the edge list we've created before
g_np <- as.network(g_elist_from_alist, matrix.type = "edgelist", vertices = nodelist)
g_np

plot(g_np)
```


# Modifying networks in igraph

```{r}
#can make changes to data once it's in a network object

#e.g. removing a node
g_mod <- delete_vertices(g,2)
plot(g_mod)

#e.g. removing all nodes with age=21
g_mod <- delete_vertices(g,which(V(g)$age == 21))
plot(g_mod)

#e.g. removing all nodes with missing age
g_mod <- delete_vertices(g,which(is.na(V(g)$age)))
plot(g_mod)

#using V() as an alternative to set_vertex_attr()
V(g)$gender <- c("male", "female", "female", "other", "male") 
g

#using E() to access, add and modify edge attributes
#??? no code

#functions that modify the entire graph
#e.g. reversing the direction of all edges
g_rev <- reverse_edges(g)
plot(g_rev)

#combining edges of two graphs
g_union <- igraph::union(g, g_rev)

#removing directions
g_undir <- as_undirected(g)
plot(g_undir)
```
