---
title: "LaLab 4: Networks as causes - Spatial regression analysis"
author: "Jola Gockel"
date: "2026-05-01"
output: html_document
---

```{r, globalsettings, echo=FALSE, warning=FALSE, results='hide'}
library(knitr)

knitr::opts_chunk$set(echo = TRUE)
opts_chunk$set(tidy.opts=list(width.cutoff=100),tidy=TRUE, warning = FALSE, message = FALSE,comment = "#>", cache=TRUE, class.source=c("test"), class.output=c("test2"))
options(width = 100)
rgl::setupKnitr()

colorize <- function(x, color) {sprintf("<span style='color: %s;'>%s</span>", color, x) }
```

```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(position = c('top', 'right'))
#klippy::klippy(color = 'darkred')
#klippy::klippy(tooltip_message = 'Click to copy', tooltip_success = 'Done')
```

```{r setup, include=FALSE}
library(spdep) #for the spatial weight matrix and for the Moran's I test
library(spatialreg) #for the spatial lag and spatial error models
library(haven)
library(tidyverse)
library(igraph)
library(lme4)
```

Last compiled on `r format(Sys.time(), '%B, %Y')`

<br>

# Constructing a Spatial Weight Matrix

```{r}
data(columbus)
mydata <- columbus
attach(mydata)

neighbors <- col.gal.nb #an adjacency list of neighbours
class(neighbors) #is already a nb object so we can skip additional steps
#transforming nb object to a listw object
listw <- nb2listw(neighbors, style = "W") #W indicates row-standardization
summary(listw)
```


# Linear Regression Analysis

```{r}
summary(CRIME)
summary(INC)
summary(HOVAL)

plot(neighbors, coords)

olsreg <- lm(CRIME ~ INC + HOVAL)
summary(olsreg)
```
Higher household income and higher housing values lead to a significantly lower crime rate.BUT the model does not take the spatial dependence into account and our hypothesis tests may lead to erroneous conclusions.


# Spatial Analysis

## Moran’s I test

```{r}
#to figure out if we even need a spatial model
moran.test(CRIME, listw)
```
The p-value is very small so we reject the null hypothesis of no spatial dependence.


## Lagrange multiplier tests for spatial lag and spatial error dependencies

```{r}
#to figure out which type of spatial models we should use
lm.LMtests(olsreg, listw, test=c("LMlag", "LMerr"))
```
Both hypotheses are rejected, so we need theory to decide whether to use a spatial lag or spatial error model (or test them all and see if results differ).

## Spatial lag model (spatial autoregressive model)

```{r}
slm <- lagsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = "FALSE")
summary(slm)
#coefficients are very similar to the LRM coefficients
#the spatial parameter rho is positive and significant showing spatial dependence

#direct and indirect effects
im<-impacts(slm, listw = listw, zstats=TRUE, R=200) #zstats=T to obtain p-values
sums<-summary(im,  zstats=T)
data.frame(sums$res)
data.frame(sums$pzmat)
#both total effects are significant but mainly driven by direct effects
```


## Spatial error model

```{r}
sem <- errorsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = FALSE)
summary(sem)
#again a significantly negative effect of household income and housing values on the crime rate
#the spatial parameter lambda is positive and significant showing dependence
```


## Spatial autoregressive combined model

```{r}
sac <- sacsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = FALSE)
summary(sac)
#similar results, but rho (and lambda?) is now not significant
```


## Spatial Durbin model

```{r}
sdm <- lagsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = TRUE)
summary(sdm)
#the results are similar and the lags are not significant
```


## Spatial Durbin error model

```{r}
sdem <- errorsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = TRUE)
summary(sdem)
#results are again similar, but now the lag of household income is significant
```


## General nested model

```{r}
gnm <- sacsarlm(CRIME ~ INC + HOVAL, data = mydata, listw, Durbin = TRUE)
summary(gnm)
#similar results again
#so choose model based on theory
```


# Assignment

## Data + Data Cleaning

```{r}
PupilsWaveV <- read_dta(file = "PupilsWaveV.dta")

PupilsWaveV$acthomew <- 5-PupilsWaveV$acthomew
PupilsWaveV$sex <- PupilsWaveV$sex-1
PupilsWaveV$sex <- 1-PupilsWaveV$sex
```


## Linear Regression Model

```{r}
olsreg <- lm(acthomew ~ sex + age, data=PupilsWaveV)
summary(olsreg)
```


## Multilevel Linear Model

```{r}
ml <- lmer(acthomew ~ sex + age + (1|schoolnr), PupilsWaveV)
summary(ml)
```


## Friends Adjacency Matrix

```{r}
class12b <- PupilsWaveV %>%
  filter(schoolnr=="12b")

#fetching the spatial weight matrix (adjacency list format)
friendsadjlist <- class12b %>%
  select(namenr, friend1:friend12)
class(friendsadjlist)

#transform into edge list
friendsedgelist <- class12b %>%
  select(namenr, friend1:friend12) %>%
  pivot_longer(c(friend1:friend12)) %>%
  filter(!is.na(value)) %>%
  rename(from="namenr", to="value", sourcevar="name") %>%
  relocate(to, .after=from)

friendsnodelist <- class12b %>%
  select(namenr, sex, age, acthomew) %>%
  rename(name="namenr")

#transform into igraph object
friendsgraph <- graph_from_data_frame(friendsedgelist, vertices=friendsnodelist)
class(friendsgraph)

#plot
plot(friendsgraph)

layout <- layout_with_kk(friendsgraph)
plot(friendsgraph, layout=layout, edge.arrow.size=.5)

#transform to adjacency matrix
friendsadjmatrix <- as_adjacency_matrix(friendsgraph, sparse=F)
class(friendsadjmatrix)
friendsadjmatrix
```


## Row-standardized Spatial Weight Matrix

```{r}
friends <- mat2listw(friendsadjmatrix, row.names=NULL, zero.policy=T, style="W")
class(friends)
```


## Moran’s I Test + Lagrange Multiplier Tests

```{r}
moran.test(class12b$acthomew, friends, zero.policy=T) #=T bc we have isolates

olsregnew <- lm(acthomew ~ sex + age, data=friendsnodelist)
summary(olsregnew)

lm.LMtests(olsregnew, friends, test=c("LMlag", "LMerr"), zero.policy=T)
```
Reject null hypothesis for LMlag but not for LMerr --> use spatial lag model

## Spatial Lag Model + Impacts

```{r}
slm <- lagsarlm(acthomew ~ sex + age, data=friendsnodelist, friends, zero.policy=T, Durbin=F)
summary(slm)

#impacts
im<-impacts(slm, listw=friends, zstats=T, R=200)
sums<-summary(im,  zstats=T)
data.frame(sums$res)
data.frame(sums$pzmat)
sums$direct
sums$indirect
sums$total
```
Ignoring significance because of low sample size. Girls tend to do more homework than boys. The older you are, the more homework you do. Negative rho: if your friends do homework, you are less likely to do it.
Impacts: Direct effects tend to be larger than indirect effects and have the reversed sign.
ADD MORE INTERPRETATION

## Spatial Error Model + Spatial Autoregressive Combined Model

```{r}
sem <- errorsarlm(acthomew ~ sex + age, data=friendsnodelist, friends, zero.policy=T, Durbin=F)
summary(sem)

sac <- sacsarlm(acthomew ~ sex + age, data=friendsnodelist, friends, zero.policy=T, Durbin=F)
summary(sac)
#add impacts
```
SEM: Can just interpret point estimates. Lambda is negative so unobserved factors of the friends are reducing your homework.
SAC: Same interpretation as above.

## Spatial Durin Model + Spatial Durbin Error Model + General Nested Model

```{r}
sdm <- lagsarlm(acthomew ~ sex + age, data=friendsnodelist, friends, zero.policy=T, Durbin=T)
summary(sdm)
#add impacts

sdem <- errorsarlm(acthomew ~ sex + age, data=friendsnodelist, friends, zero.policy=T, Durbin=T)
summary(sdem)

gnm <- sacsarlm(acthomew ~ sex + age, data=friendsnodelist, friends, zero.policy=T, Durbin=T)
summary(gnm)
#add impacts
```
ADD INTERPRETATION

## Conclusion
Think*
